---
title: "Lab 6"
author: "Karan Nayak"
date: "10/21/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/UMich_Bio201_F19/")
```

# Load packages
```{r Load packages, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
library(agricolae)
set.seed(7)
```

# Import data

### Import + format SCFA data

Import the SCFA data frame from last week. Update the columns names to match the course's style, set the order of the semesters by converting from character to factor. 

```{r include=FALSE}
scfa_long <- read_delim("~/Documents/UMich_Bio201_F19/Lab5/curated_data/SCFA_wkly_long.txt", 
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("NA")) %>%
  rename_all(tolower) %>%
  select(-ends_with("median")) %>% 
  mutate(totalscfa_mean = acetate_mean + butyrate_mean + propionate_mean,
         semester = factor(semester,
                           levels=c("Winter2015", "Fall2015", 
                                    "Winter2016", "Fall2016", 
                                    "Winter2017", "Fall2017", 
                                    "Winter2018", "Fall2018", 
                                    "Winter2019"), ordered = TRUE)) %>%
  filter(quantity_compliant == "yes")
```

### Import tax data 

Import the taxonomic table `person_week_mean_by_cat_w_phylum.csv`. Notice the extension is csv, the delimitor in this file is a comma instead of a tab (which is different from what we have worked with previously). 
```{r}
# taxanomic data
taxa_category <- read_delim("~/Documents/UMich_Bio201_F19/Lab6/raw_data/person_week_mean_by_cat_w_phylum.csv",
                            delim = ",", escape_double = FALSE, 
                            trim_ws = TRUE, na=c("NA"),
                            col_types = list())
View(taxa_category)
```

Look at the data frame, notice the week and semester formats are different then the rest of our data files. Today we are going to do some correlations between taxon abundance and short chain fatty acid (SCFA) concentrations. If we want to join these two data sets we will need make sure the format of these variables are the same. 

### Format tax data
```{r}
# create duplicate of data frame
taxa_cat_curated <- taxa_category %>%
  rename_all(tolower) %>%
  rename(participant_id = person,
         study_week = week,
         relative_abundance = fract) %>%
  select(participant_id, study_week, semester, tax_id, relative_abundance, phylum, tax_name, category)

# update week format 
taxa_cat_curated[["study_week"]] <- recode(.x=taxa_cat_curated[["study_week"]], 
                                  "week_1"="week1")
taxa_cat_curated[["study_week"]] <- recode(.x=taxa_cat_curated[["study_week"]], 
                                  "week_3"="week3")
# update semester format
taxa_cat_curated[["semester"]] <- recode(.x=taxa_cat_curated[["semester"]], 
                                  "2015w"="Winter2015")
taxa_cat_curated[["semester"]] <- recode(.x=taxa_cat_curated[["semester"]], 
                                  "2015f"="Fall2015")
taxa_cat_curated[["semester"]] <- recode(.x=taxa_cat_curated[["semester"]], 
                                  "2016w"="Winter2016")
taxa_cat_curated[["semester"]] <- recode(.x=taxa_cat_curated[["semester"]], 
                                  "2016f"="Fall2016")
taxa_cat_curated[["semester"]] <- recode(.x=taxa_cat_curated[["semester"]], 
                                  "2017w"="Winter2017")
taxa_cat_curated[["semester"]] <- recode(.x=taxa_cat_curated[["semester"]], 
                                  "2017f"="Fall2017")
taxa_cat_curated[["semester"]] <- recode(.x=taxa_cat_curated[["semester"]], 
                                  "2018w"="Winter2018")
taxa_cat_curated[["semester"]] <- recode(.x=taxa_cat_curated[["semester"]], 
                                  "2018f"="Fall2018")
taxa_cat_curated[["semester"]] <- recode(.x=taxa_cat_curated[["semester"]], 
                                  "2019w"="Winter2019")

# Save a copy of the curated data frame
write_delim(taxa_cat_curated, path="Lab6/curated_data/taxa_cat_curated.csv",
            delim = ",", quote = FALSE, col_names = TRUE)

View(taxa_cat_curated)
```

The cleaned-up `taxa_cat_curated` contains a subset of organisms found in the gut, specifically organisms which consume or produce metabolites of interest. The `tax_id` column is a number assigned to a sequence during the sequence processing pipeline. You won't need to use these right now. The `relative_abundance` column contains the frequencey of the organisms in each participant. The frequency is the weekly mean of (# sequences for each tax_id) / (total sequences per sample) for each participant. The `tax_name` is the most likely identity of each sequence. Information regarding microbial metabolism is in the `category` column:

* RS_degraders = Resistant starch degrading specialists (Bifidobacterium + Ruminococcus bromii)
* butyrogen_focused = A focused list of the most common / canonical butyrogens
* butyrogen_flint = A broader list of butyrogenic organisms (information comes from Harry Flint's research)
* hydrogen_consumers = Hydrogen consumers
* hydrogen_gen = Hydrogen generators
* propionate_gen = Propionate generators (essentially Bacteroides)

### Join SCFA + tax data

In the latter half of this lab we are going to test correlations. To test correlations between taxa and their metabolites, we need to join the data frames containing these data. Join the `tax_cat_curated` and `scfa_long` data frames. Remember to check the number of rows and participants in each data frame to help you select the best type of join. Discuss your decision with your neighbor, did you choose the same type of join? 

```{r}
#check number of rows
nrow(taxa_cat_curated) 
nrow(scfa_long) 

#check number of participants 
n_distinct(taxa_cat_curated$participant_id) 
n_distinct(scfa_long$participant_id) 

#identify shared columns = ?
colnames(taxa_cat_curated) 
colnames(scfa_long)
```

```{r}
tax_scfa <- inner_join(scfa_long, taxa_cat_curated,
                 by = c())
```

# Intro

Last week we discussed the factors influencing data structure (sample size, varaince), how to check assumptions for statistical tests, how to determine if the differences between two means (t-tests, Mann-Whitney-U) or more (ANOVA) are likely real, and what to do if your datset does not meet the assumptions when comparing two means (non-parametric). 

In the first portion of the lab we will review the assumption checking process for comparison of more than 2 means, use the non-pararmetric counterpart to ANOVA (Kruskall-Wallace), and learn what to do if your ANOVA (or Kruskall-Wallace) test has a significant result. This will complete our section on categorical/discrete vs. continuous variables. The second half of the lab will discuss how to plot and analyze continuous vs. countinuous variables.

### Checking assumptions

Remember the assumptions for an ANOVA are similar to a t-test (large-ish sample size, normal distribution, equal variances). 

```{r eval=FALSE, include=FALSE}
# sample size of each group
group_by() %>% summarize(sample_size = n())

# check normality of each group
shaprio.test()
histogram()

# check variances
bartlett.test()
```

Last week each ANOVA you conducted should have resulted in a non-significant result (approx p >= 0.15). Run the code for the plot and ANOVA below. 

```{r}
# know assumptions are already met
scfa_long %>%
  filter(study_week == "week1") %>%
  ggplot(aes(x = semester,
             y = butyrate_mean, 
             color = semester)) +
  geom_violin() + 
  geom_jitter() + 
  xlab(NULL) +
  ylab("Week 1 butyrate (mmol/kg)") + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# ANOVA 
aov_wk1 <- aov(butyrate_mean ~ semester, scfa_long)
summary(aov_wk1)
```

Here we have a small p-value, this means at least one of our groups is significantly different from the others. To identify which group is different we have to run a pos-hoc test (an additional test). We will use Tukey's Honestly Significant Difference (HSD).

```{r}
# conduct Tukey's HSD test 
tuk_wk1 <- tidy(TukeyHSD(aov_wk1)) %>%
  mutate(adj.p.value = round(adj.p.value, digits = 2)) %>%
  filter(adj.p.value <= 0.10) #filter Tukey output for significant results 

# save data frame of results to file 
write_delim(tuk_wk1, path = "~/Documents/UMich_Bio201_F19/Lab6/curated_data/tuk_wk1.txt",
            delim = "\t", col_names = TRUE, quote = FALSE)
```

### Kruskall-Wallace

If you have a data set that is clearly non-normal (such as the breath methane data), and want to ask a similar question as we asked with ANOVAs last week (e.g., Does the breath methane differ during week 3 for individuals consuming different supplements?), we need to use a non-parametric test. The non-parametric companion to an ANOVA is the Kruskall-Wallace test. 

Last week you should have checked assumptions on breath methane and determined that it was non-normal, so an ANOVA would not be an appropriate test to use to determine if the week 3 breath methane concentrations differed between supplements consumed.

```{r}
# import methane data 
breath_wkly <- read_delim("~/Documents/UMich_Bio201_F19/Lab5/curated_data/breath_wkly.txt", 
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("NA")) %>%
# format methane data
  rename_all(tolower) %>%
  filter(study_week == "week3",
         quantity_compliant == "yes") %>% #remove non-compliant individuals
  mutate(semester = factor(semester,
                           levels=c("Winter2015", "Fall2015", 
                                    "Winter2016", "Fall2016", 
                                    "Winter2017", "Fall2017", 
                                    "Winter2018", "Fall2018", 
                                    "Winter2019"), ordered = TRUE))
# plot
breath_wkly %>%
  ggplot(aes(x = supplement_consumed, 
             y = ch4_mean, 
             color = supplement_consumed)) + 
  geom_violin() + 
  geom_jitter() + 
  xlab(NULL) + 
  ylab("Breath methane (ppm)") + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# run the test
kruskal.test(supplement_consumed ~ ch4_mean, breath_wkly)
```

The interpretation for Kruskall-Wallace is similar to an ANOVA; small p-value indicates at least one group is different from the others. 

### Homework 6.1

* Write hypotheses to answer the question; do hydrogen breath levels differ between participants consuming fiber supplements? Create a plot of hydrogen levels:
```{r}
# Hypothesis: 
#If more fiber supplements are consumed, hydrogen breath levels will increase because of increased fermentation and production.

# plot 
breath_wkly %>%
  ggplot(aes(x = supplement_consumed, 
             y = h2_mean, 
             color = supplement_consumed)) + 
  geom_violin() + 
  geom_jitter() + 
  xlab(NULL) + 
  ylab("Hydrogen (ppm)") + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

* Which assumptions need checked for the statistical test(s)? Check the assumptions in the space below:
```{r}
# check sample size
breath_wkly %>%
  group_by(supplement_consumed) %>% 
  summarise(sample_size = n()) 

# normal distribution 
h2_df1 <- breath_wkly %>%
  filter(supplement_consumed == "Accessible")
shapiro.test(h2_df1$h2_mean) #Normal

h2_df2 <- breath_wkly %>%
  filter(supplement_consumed == "Arabino") 
shapiro.test(h2_df2$h2_mean) #Normal

h2_df3 <- breath_wkly %>%
  filter(supplement_consumed == "BRMPS")
shapiro.test(h2_df3$h2_mean) #Not Normal

h2_df4 <- breath_wkly %>%
  filter(supplement_consumed == "BRMPS+Accessible")
shapiro.test(h2_df4$h2_mean) #Normal

h2_df5 <- breath_wkly %>%
   filter(supplement_consumed == "HiMaize") 
shapiro.test(h2_df5$h2_mean) #Not Normal

h2_df6 <- breath_wkly %>%
  filter(supplement_consumed == "HiMaize+BRMPS")
shapiro.test(h2_df6$h2_mean) #Not Normal

h2_df7 <- breath_wkly %>%
  filter(supplement_consumed == "Inulin") 
shapiro.test(h2_df7$h2_mean) #Not Normal

h2_df8 <- breath_wkly %>%
  filter(supplement_consumed == "LOODAT") 
shapiro.test(h2_df8$h2_mean) #Not Normal

h2_df10 <- breath_wkly %>%
 filter(supplement_consumed == "Psyllium") 
shapiro.test(h2_df10$h2_mean) #Normal

h2_df11 <- breath_wkly %>%
  filter(supplement_consumed == "Psyllium+BRMPS") 
shapiro.test(h2_df11$h2_mean) #Normal

h2_df12 <- breath_wkly %>%
  filter(supplement_consumed == "transition_HiMaize")
shapiro.test(h2_df12$h2_mean) #Not Normal
 
#p value is 0.05

ggplot(h2_df, aes(x=h2_mean)) +
  geom_histogram() 
qqnorm(h2_df$h2_mean); qqline(h2_df$h2_mean)


# variances
bartlett.test(h2_mean ~ supplement_consumed, data = breath_wkly)
```

* Which statistical test(s) would you use to answer the question? Conduct the appropriate statistical test(s):
```{r}
# Some supplements are non-normal whereas others are normal. There are unequal variances between the groups. 

kruskal.test(supplement_consumed ~ h2_mean, breath_wkly)
```

* What is the answer to the question? Write your conclusions:
```{r}
# conclusion 

#Since the p-value is large, it indicates that no groups are significantly different from the others. 

```

# Correlations

Correlations are used to determine if there is a relationship between two continuous variables. For the questions we are interested in, determining if there is a monotonic (or linear) relationship will provide insight. A monotonic relationship would occur when both continuous variables increase or decrease simultaneously, or when one variable increases the other decreases. 

To test correlations in this course, we are going to use the best practice approach of linear models (lm). However we do want to point out other correlation tests in case you encounter them elsewhere (Pearson which is parametric and Spearman Rank with is non-parametric). These tests have assumptions which are easy to violate, and the statistics provided (p and R values) do not have a straightfoward interpretation.

### Plot 2 continuous variables

To plot two continuous variables we will use a combination of geom_point() and geom_smooth(). 

```{r}
tax_scfa %>%
  filter(category == "butyrogen_focused",
         study_week == "week3",
         supplement_consumed == "HiMaize") %>%
  ggplot(aes(x = butyrate_mean,
             y = relative_abundance)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
  xlab("Week 3 mean butyrate (mmol/kg)") + 
  ylab("Relative abundance") 
```

You plot two continuous variables and they appear to be correlated, you can determine if they are correlated by building a linear model. 

### Linear models

The linear model function is lm(). The linear models we will use have similar arguments as running an ANOVA. You must have the two variables (x and y) and a data frame in which each variable is a column. To access the results of the model, assign it to a new object. In the example below we will determine if there is a correlation between butyrate concentration and the abundance of some butyrate producing organisms. 

```{r}
m1 <- tax_scfa %>%
  filter(category == "butyrogen_focused", #subset dataframe to butyrate producers
         study_week == "week3", #subset for week3
         supplement_consumed == "HiMaize") %>% #subset for HiMaize consumers 
  lm(butyrate_mean ~ relative_abundance, data = .) #test relationship
summary(m1) #view results 
```

Look at the p-value in the same line as the F-statistic, this tells us if there is a significant result. This p-value is high (p-value = 0.2959), so we do not have significance. However if it was significant, we would then look at the adjusted R-squared value which indicates the quality of linear fit (higher R-squared is better >=0.8).

### Homework 6.2

Use the `tax_scfa` data frame, calculate the ratio of each SCFA in a new column (scfa/total). Check that your ratios sum to 1. 

```{r}
scfa_ratio <- tax_scfa %>%
  mutate(acetate_ratio = acetate_mean/totalscfa_mean) %>% 
  mutate(butyrate_ratio = butyrate_mean/totalscfa_mean) %>% 
  mutate(propionate_ratio = propionate_mean/totalscfa_mean) %>%
  mutate(ratio_check = acetate_ratio + butyrate_ratio + propionate_ratio)
```

Calculate the ratios of Firmicutes to Bacteroidetes (Firmicutes abund / Bacteroidetes abundance).

```{r}
# filter for Firmicutes, calculate sum per person per week (firmicutes_abund)
Fr <- scfa_ratio %>%
 filter(phylum == "Firmicutes") %>%
 group_by(participant_id, study_week) %>%
 summarize(firmicutes_abund = sum(relative_abundance))
# filter for Bacteroidetes, calculate sum per person per week (bacteroidetes_abund)
Br <- scfa_ratio %>%
 filter(phylum == "Bacteroidetes <phylum>") %>%
 group_by(participant_id, study_week) %>%
 summarize(bacteroidetes_abund = sum(relative_abundance))
# join subsetted data frames
tax_ratio <- inner_join(Fr, Br) %>%
  mutate(f_b_ratio = (firmicutes_abund / bacteroidetes_abund)) #calculate ratios
  inner_join(tax_ratio, scfa_ratio) #join with scfa_ratio data frame
tax_ratio1 <- inner_join(tax_ratio, scfa_ratio)
```

### Homework 6.3
Determine if there is a relationship between the butyrate concentration and Firmicutes abundance. Conduct the statistical test and generate a plot. Save plot to folder within lab 6 called 'figures'.
```{r}
# plot
  plot_tax1 <- tax_ratio1 %>%
  ggplot(aes(x = butyrate_mean,
             y = firmicutes_abund)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
  xlab("Mean butyrate (mmol/kg)") + 
  ylab("Firmicutes abundance") 

save_plot(filename = "~/Documents/UMich_Bio201_F19/Lab6/Figures/ButyrateFirmicutes.pdf",
          plot = plot_tax1,
          nrow = 3, ncol = 6,
          base_aspect_ratio = 1.1)

# correlation test 
m63a <- tax_ratio1 %>%
  lm(butyrate_mean ~ firmicutes_abund, data = .) 
summary(m63a)

##The p value is less than 2.2e-16 so the data is significant. The r squared value is low (0.003863) so the two variables are not very correlated.
```

Do you draw the same conclusions if you use the butyrate ratio and Firmicutes to Bacteroidetes ratio (f_b_ratio)?
```{r}
# plot
  plot_tax2 <- tax_ratio1 %>%
  ggplot(aes(x = butyrate_ratio,
             y = f_b_ratio)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
  xlab("Butyrate Ratio (mmol/kg)") + 
  ylab("Firmicutes to Bacteroidetes Ratio") 

save_plot(filename = "~/Documents/UMich_Bio201_F19/Lab6/Figures/FirmicutesToBacteroidetes.pdf",
          plot = plot_tax2,
          nrow = 3, ncol = 6,
          base_aspect_ratio = 1.1)

# correlation test 
m63b <- tax_ratio1 %>%
  lm(butyrate_ratio ~ f_b_ratio, data = .) 
summary(m63b)

#The results are the same as the relationship between the butyrate concentration and Firmicutes abundance. The p value is less than 2.2e-16 so the data is significant. The r squared value is low (0.02854) so the two variables are not very correlated.
```

### Homework 6.4
Determine if there is a relationship between the propionate concentration and Bacteroidetes abundance. Conduct the statistical test and generate a plot. Save the plot.
```{r}
# plot
  plot_tax3 <- tax_ratio1 %>%
  ggplot(aes(x = propionate_mean,
             y = bacteroidetes_abund)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
  xlab("Mean Propionate (mmol/kg)") + 
  ylab("Bacteroidetes Abundance") 

save_plot(filename = "~/Documents/UMich_Bio201_F19/Lab6/Figures/PropionateBacteroidetes.pdf",
          plot = plot_tax3,
          nrow = 3, ncol = 6,
          base_aspect_ratio = 1.1)

# correlation test 
m63c <- tax_ratio1 %>%
  lm(propionate_mean ~ bacteroidetes_abund, data = .) 
summary(m63c)

#The p value is less than 2.2e-16 so the data is significant. The r squared value is low (0.04197) so the two variables are not very correlated.
```

Do you draw the same conclusions if you use the propionate ratio and Firmicutes to Bacteroidetes ratio?
```{r}
# plot
  plot_tax4 <- tax_ratio1 %>%
  ggplot(aes(x = propionate_ratio,
             y = f_b_ratio)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
  xlab("Propionate Ratio (mmol/kg)") + 
  ylab("Firmicutes to Bacteroidetes Ratio") 

save_plot(filename = "~/Documents/UMich_Bio201_F19/Lab6/Figures/PropionateFirmicutesToBacteroidetes.pdf",
          plot = plot_tax4,
          nrow = 3, ncol = 6,
          base_aspect_ratio = 1.1)

# correlation test 
m63d <- tax_ratio1 %>%
  lm(propionate_ratio ~ f_b_ratio, data = .) 
summary(m63d)

#The results are the same as the relationship between the propionate concentration and Bacteroidetes abundance. The p value is less than 2.2e-16 so the data is significant. The r squared value is low (0.02136) so the two variables are not very correlated.
```

### Homework 6.5
Import the curated pH+BSS data you generated during Lab4. Join the pH data frame with the tax_ratio data frame.
```{r}
# import 
pH_BSS_wkly <- read_delim("~/Documents/UMich_Bio201_F19/Lab5/curated_data/pH_wkly.txt", 
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("NA")) %>%
  rename_all(tolower)

# join 
tax_scfa_pH <- inner_join(tax_ratio1, pH_BSS_wkly,
                 by = c()) %>%
   na.omit(semester)

```

Determine if there is a relationship between total SCFAs and acetate. Conduct the statistical test and generate a plot. Save the plot to figures folder.
```{r}
# plot
  plot_tax5 <- tax_scfa_pH %>%
  ggplot(aes(x = totalscfa_mean,
             y = acetate_mean)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
  xlab("Mean Total Scfa (mmol/kg)") + 
  ylab("Mean Acetate") 

save_plot(filename = "~/Documents/UMich_Bio201_F19/Lab6/Figures/AcetateTotalSCFA.pdf",
          plot = plot_tax5,
          nrow = 3, ncol = 6,
          base_aspect_ratio = 1.1)

# correlation test 
m63e <- tax_scfa_pH %>%
  lm(totalscfa_mean ~ acetate_mean, data = .) 
summary(m63e)

#The p value is less than 2.2e-16 so the data is significant. The r squared value is high (0.8828) so the two variables are pretty related.

```

Do you draw the same conclusions if you use the total SCFAs and acetate ratio?
```{r}
# plot
  plot_tax6 <- tax_scfa_pH %>%
  ggplot(aes(x = totalscfa_mean,
             y = acetate_ratio)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
  xlab("Mean Total Scfa (mmol/kg)") + 
  ylab("Acetate Ratio (mmol/kg)") 

save_plot(filename = "~/Documents/UMich_Bio201_F19/Lab6/Figures/AcetateRatioTotalSCFA.pdf",
          plot = plot_tax6,
          nrow = 3, ncol = 6,
          base_aspect_ratio = 1.1)

# correlation test 
m63f <- tax_scfa_pH %>%
  lm(totalscfa_mean ~ acetate_ratio, data = .) 
summary(m63f)

# The same conclusions are not drawn, but the p value 0.003033 is low so the data is significant. The r squared value is low (0.0009486) so the two variables are not very correlated.
```

### Homework 6.6
Determine if there is a relationship between total SCFAs and pH. Conduct the statistical test and generate a plot. Save the plot.
```{r}
# plot
  plot_tax7 <- tax_scfa_pH %>%
  ggplot(aes(x = totalscfa_mean,
             y = ph_mean)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
  xlab("Mean Total Scfa (mmol/kg)") + 
  ylab("Mean pH (mmol/kg)") 

save_plot(filename = "~/Documents/UMich_Bio201_F19/Lab6/Figures/TotalSCFApH.pdf",
          plot = plot_tax7,
          nrow = 3, ncol = 6,
          base_aspect_ratio = 1.1)

# correlation test 
m63g <- tax_scfa_pH %>%
  lm(totalscfa_mean ~ ph_mean, data = .) 
summary(m63g)

#The p value is less than 2.2e-16 so the data is significant. The r squared value (0.03747) is low so the two variables are not very correlated.
```

### Homework 6.7
Determine if there is a relationship between butyrate and pH. Conduct the statistical test and generate a plot. Save the plot.
```{r}
# plot
  plot_tax8 <- tax_scfa_pH %>%
  ggplot(aes(x = butyrate_mean,
             y = ph_mean)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
  xlab("Mean Butyrate (mmol/kg)") + 
  ylab("Mean pH (mmol/kg)") 

save_plot(filename = "~/Documents/UMich_Bio201_F19/Lab6/Figures/ButyratepH.pdf",
          plot = plot_tax8,
          nrow = 3, ncol = 6,
          base_aspect_ratio = 1.1)

# correlation test 
m63h <- tax_scfa_pH %>%
  lm(butyrate_mean ~ ph_mean, data = .) 
summary(m63h)

#The p value is less than 2.2e-16 so the data is significant. The r squared value (0.04633) is low so the two variables are not very correlated.
```

### Homework 6.8
Save each data frame generated today:

* taxa_cat_curated
* scfa_ratio
* tax_scfa
* tax_scfa_pH

```{r}
write_delim(taxa_cat_curated, 
            path = "~/Documents/UMich_Bio201_F19/Lab6/curated_data/taxa_cat_curated.txt",
            delim = "\t", quote = FALSE, col_names = TRUE)

write_delim(scfa_ratio, 
            path = "~/Documents/UMich_Bio201_F19/Lab6/curated_data/scfa_ratio.txt",
            delim = "\t", quote = FALSE, col_names = TRUE)

write_delim(tax_scfa, 
            path = "~/Documents/UMich_Bio201_F19/Lab6/curated_data/tax_scfa.txt",
            delim = "\t", quote = FALSE, col_names = TRUE)

write_delim(tax_scfa_pH, 
            path = "~/Documents/UMich_Bio201_F19/Lab6/curated_data/tax_scfa_pH.txt",
            delim = "\t", quote = FALSE, col_names = TRUE)
```

-----
end